---
title: 'Predicting Violent Incidents in New Orleans: An Analysis of Calls for Service Data'
author: "c-burruss"
date: "March 14, 2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# 1. Introduction and Overview

## 1.1 Executive Summary

This analysis explores the relationship between various urban disorder indicators and violent incidents in New Orleans neighborhoods. By examining data from the city's calls for service (CFS) system alongside other civic datasets, we investigate whether factors such as code violations, streetlight outages, abandoned vehicles, and lot abatements can serve as predictors of violent incidents. Understanding these relationships can assist city planners, law enforcement, and community organizations in prioritizing resources and interventions to improve public safety.

## 1.2 Dataset Description

The analysis utilizes several datasets from the City of New Orleans open data portal:

1. **Calls for Service (2023)**: Police response data including incident type, location, and time.
2. **Code Violations**: Records of property maintenance and zoning violations.
3. **311 Calls**: Citizen reports of issues including streetlight outages and abandoned vehicles.
4. **Chapter 66 Lot Abatements**: Records of city-initiated lot clean-ups for properties in violation.
5. **Neighborhood Statistical Areas**: Geographic boundaries defining New Orleans neighborhoods.

## 1.3 Project Goals

The primary objectives of this analysis are to:

1. Identify spatial patterns of violent incidents across New Orleans neighborhoods.
2. Assess correlations between urban disorder indicators and violent incidents.
3. Develop predictive models to determine which factors best predict neighborhood violence.
4. Provide data-driven insights to inform community safety initiatives.

# 2. Methods and Analysis

## 2.1 Data Preparation

### 2.1.1 Setting Up the Environment

```{r initialization, results=FALSE}
# Machine learning packages
packages_ml <- c('tidymodels', 'glmnet', 'ranger', 'earth', 'vip')
# Spatial analysis packages
packages_spatial <- c('tidygeocoder', 'geojsonio', 'spdep', 'sf')
# Data wrangling and viz packages
packages_viz <- c('tidyverse', 'ggplot2', 'ggthemes', 'ggrepel',
                 'janitor', 'knitr', 'readr', 'grid')
# API packages
packages_api <- c('RSocrata')

# Function to check and install packages
install_if_missing <- function(package) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package, repos = 'source')
    library(package, character.only = TRUE)
  }
}

# Apply to all categories
lapply(c(packages_ml, packages_spatial, packages_viz, packages_api), 
       install_if_missing)

# Ensure that dplyr's filter() function works
conflict_prefer('filter', 'dplyr')
```

### 2.1.2 Helper Functions

```{r helper_functions}
# Helper function to clean and standardize data
pasteurize <- function(df) {
  # Helper function to clean column names
  clean_column_names <- function(col_name) {
    cleaned <- tolower(col_name)
    cleaned <- gsub("[^a-z0-9_]", "_", cleaned)
    cleaned <- gsub("_+", "_", cleaned)
    cleaned <- gsub("^_|_$", "", cleaned)
    return(cleaned)
  }
  
  # Helper function for title case conversion
  to_title_case <- function(x) {
    # Ensure we're working with a character vector
    if (!is.character(x)) return(x)
    # Apply transformation to each element in the vector
    sapply(x, function(str) {
      # Handle NA values
      if (is.na(str)) return(NA)
      # Split into words and capitalize first letter of each
      words <- strsplit(tolower(str), " ")[[1]]
      words <- paste0(toupper(substr(words, 1, 1)),
                     substr(words, 2, nchar(words)))
      # Rejoin words with spaces
      paste(words, collapse = " ")
    })
  }
  
  # Main data cleaning operations
  df %>%
    filter(!if_all(everything(), is.na)) %>%
    distinct() %>%
    rename_with(clean_column_names) %>%
    mutate(across(where(is.character),
                 ~str_trim(.) %>%
                   to_title_case))
}

# Function for summarizing dataset statistics
stat.summarize <- function(df) {
  numeric_df <- df %>%
    select(where(is.numeric))
  
  # Initialize a list to store results
  summary_list <- list()
  
  # Loop through each numeric column and calculate statistics
  for (col in names(numeric_df)) {
    summary_list[[col]] <- c(
      min = min(numeric_df[[col]], na.rm = TRUE),
      max = max(numeric_df[[col]], na.rm = TRUE),
      median = median(numeric_df[[col]], na.rm = TRUE),
      mean = mean(numeric_df[[col]], na.rm = TRUE),
      sd = round(sd(numeric_df[[col]], na.rm = TRUE), 2),
      n = sum(!is.na(numeric_df[[col]]))
    )
  }
  
  # Combine the results into a data frame and transpose
  result <- as.data.frame(do.call(rbind, summary_list))
  
  # Transpose the result to switch rows and columns
  result <- as.data.frame(t(result))
  
  # Set the column names to the measurements
  colnames(result) <- names(numeric_df)
  
  result %>%
    kable() %>%
    return()
}
```

### 2.1.3 Data Acquisition

```{r data_acquisition, echo = TRUE}
# Download neighborhoods spatial file
path <- './data/neighborhoods.geojson'
github_url <- 'https://raw.githubusercontent.com/CBurruss/HarvardX-CYO/main/data/neighborhoods.geojson'
if (!file.exists(path)) {
  download.file(github_url, path, mode = 'wb') # 'wb' ensures binary mode for non-text files
} else {
  message('The neighborhoods file exists!')
}

# Download 2023 calls for service (cfs)
path <- './data/cfs-2023.csv'
if(!file.exists(path)) {
  # Download data and save it to file
  cfs_file <- read.socrata('https://data.nola.gov/api/odata/v4/pc5d-tvaw')
  write.csv(cfs_file, file = path, row.names = FALSE)
  message('The cfs file has been downloaded and saved to ', path)
} else {
  message('The cfs file already exists at ', path, '!')
}

# Download code violations
path <- './data/violations.csv'
if(!file.exists(path)) {
  # Download data and save it to file
  violations_file <- read.socrata('https://data.nola.gov/api/odata/v4/3ehi-je3s')
  write.csv(violations_file, file = path, row.names = FALSE)
  message('The violations file has been downloaded and saved to ', path)
} else {
  message('The violations file already exists at ', path, '!')
}

# Download 311 calls (for streetlight outages and abandoned vehicles)
path <- './data/311.csv'
if(!file.exists(path)) {
  # Download data and save it to file
  three11_file <- read.socrata('https://data.nola.gov/api/odata/v4/2jgv-pqrq')
  write.csv(three11_file, file = path, row.names = FALSE)
  message('The 311 file has been downloaded and saved to ', path)
} else {
  message('The 311 file already exists at ', path, '!')
}

# Download ch 66 lot abatements
path <- './data/ch66.csv'
if(!file.exists(path)) {
  # Download data and save it to file
  ch66_file <- read.socrata('https://data.nola.gov/api/odata/v4/xhih-vxs6')
  write.csv(ch66_file, file = path, row.names = FALSE)
  message('The ch 66 file has been downloaded and saved to ', path)
} else {
  message('The ch 66 file already exists at ', path, '!')
}
```

## 2.2 Data Processing and Exploration

### 2.2.1 Processing Calls for Service Data

```{r process_cfs}
# Function to load and process CFS data
load_cfs_data <- function() {
  # Read in CFS data
  cfs <- read.csv("./data/cfs-2023.csv") %>%
    # Perform a general cleanse
    pasteurize() %>%
    # Convert date fields
    mutate(date = as.Date(timecreate), .after = timecreate) %>%
    mutate(year = substr(date, 1, 4), .after = date)
  
  # Extract coordinates
  cfs <- cfs %>%
    mutate(
      long_lat = str_extract(location, "\\-?\\d+\\.\\d+ \\-?\\d+\\.\\d+"),
      long_lat = str_replace(long_lat, " ", ",")
    ) %>%
    separate(long_lat, into = c("x", "y"), sep = ",") %>%
    mutate(x = as.numeric(x), y = as.numeric(y)) %>%
    na.omit()
  
  # Convert to spatial object
  cfs_sf <- cfs %>%
    st_as_sf(coords = c("x", "y"), crs = 4326, agr = "constant")
  
  return(cfs_sf)
}

cfs <- load_cfs_data()

# Display summary of the dataset
glimpse(cfs)
```

### 2.2.2 Identifying Violent Incidents

```{r identify_violent}
identify_violent_calls <- function(cfs_data) {
  # Define keywords associated with violent incidents
  violent_keywords <- c("assault", "shooting", "shot", "homicide", "weapon",
                       "burglary", "robbery", "carjack", "domestic", "murder",
                       "stab", "violence", "violent", "fight", "battery", "rape")
  
  exclusion_keywords <- c("alarm", "not occupied", "brandish", "shots fired")
  
  # Create a new column identifying violent calls
  cfs_data %>%
    mutate(
      typetext_lower = tolower(typetext),
      is_violent = str_detect(typetext_lower, paste(violent_keywords, collapse = "|")) &
        !str_detect(typetext_lower, paste(exclusion_keywords, collapse = "|")),
      predicted_category = if_else(is_violent, "Violent", "Non-violent")
    )
}

cfs <- identify_violent_calls(cfs)
violent_calls <- cfs %>%
  filter(is_violent == TRUE)

# Summarize violent call types
cfs %>%
  filter(is_violent == TRUE) %>%
  group_by(typetext) %>%
  summarize(count = n()) %>%
  mutate(percent = paste0(round(count / sum(count) * 100, 0), '%')) %>%
  arrange(desc(count)) %>%
  as_tibble() %>%
  select(-geometry) %>%
  head(15) %>%
  kable(caption = "Top 15 Violent Incident Types in New Orleans (2023)")
```

### 2.2.3 Processing Neighborhood Data

```{r process_neighborhoods}
load_neighborhoods <- function() {
  neighborhoods <- st_read('./data/neighborhoods.geojson') %>%
    clean_names() %>%
    st_transform(crs = 4326)
  
  return(neighborhoods)
}

neighborhoods <- load_neighborhoods()
```

### 2.2.4 Visualizing Neighborhoods and Violent Incidents

```{r visualize_neighborhoods, fig.cap="New Orleans Neighborhoods"}
# Simple map of neighborhoods
neighborhood_map <- ggplot() +
  geom_sf(data = neighborhoods, fill = "lightblue", color = "white") +
  labs(title = "New Orleans neighborhoods") +
  theme_solarized()

neighborhood_map
```

```{r visualize_violent_calls, fig.cap="Violent Calls for Service in New Orleans (2023)"}
# Map with violent calls as points
violent_calls_map <- ggplot() +
  geom_sf(data = neighborhoods, fill = "lightblue", alpha = 0.5, color = "white") +
  geom_sf(data = violent_calls, color = "darkgrey", alpha = 0.5, size = 0.5) +
  labs(title = "Violent calls for service in New Orleans (2023)") +
  theme_solarized()

violent_calls_map
```

### 2.2.5 Processing Risk Factor Data

```{r risk_factor_functions}
# Load code violations
load_violations <- function() {
  # Check if the processed file already exists
  if (!file.exists('./data/violations_sf.RDS')) {
    # Initial import and clean
    violations <- read.csv('./data/violations.csv') %>%
      clean_names() %>%
      mutate(violationdate = as.Date(violationdate)) %>%
      mutate(year = (substr(violationdate, 1, 4))) %>%
      filter(year == '2023') %>%
      select(-lastupload)
    
    # Standardize addresses
    violations$location_clean <- violations$location
    violations$location_clean <- gsub(',', ' and', violations$location_clean)
    violations$address <- paste(violations$location_clean, 'New Orleans, LA')
    
    # Geocode if not already
    if("lat" %in% colnames(violations) && "long" %in% colnames(violations)) {
      # Use existing coordinates
      violations_sf <- violations %>%
        st_as_sf(coords = c("long", "lat"), crs = 4326, agr = "constant") %>%
        mutate(Legend = 'Code violations') %>%
        na.omit()
    } else {
      # Geocode addresses
      violations_geo <- violations %>%
        geocode(address = address,
                method = 'arcgis',
                lat = latitude,
                long = longitude)
      
      # Convert to sf object
      violations_sf <- violations_geo %>%
        filter(!is.na(latitude) & !is.na(longitude)) %>%
        st_as_sf(coords = c("longitude", "latitude"), crs = 4326, agr = "constant") %>%
        mutate(Legend = 'Code violations')
    }
    
    # Save the processed data to an RDS file for future use
    saveRDS(violations_sf, './data/violations_sf.RDS')
    message("Geocoding complete!")
  } else {
    # Load the pre-processed data from the saved RDS
    violations_sf <- readRDS('./data/violations_sf.RDS')
    message("Loaded cached file.")
  }
  
  return(violations_sf)
}

# Load streetlight outages from 311 data
load_streetlights <- function() {
  # Check if 311.csv exists
  if(file.exists('./data/311.csv')) {
    lights <- read.csv('./data/311.csv') %>%
      clean_names() %>%
      filter(grepl('Streetlight', request_reason))
    
    # Convert to spatial object
    lights_sf <- lights %>%
      st_as_sf(coords = c('longitude', 'latitude'),
               crs = 4326,
               agr = 'constant',
               remove = FALSE) %>%
      na.omit()
    
    return(lights_sf)
  } else {
    stop("311 file not found")
  }
}

# Load abandoned vehicles from 311 data
load_vehicles <- function() {
  # Check if 311.csv exists
  if(file.exists('./data/311.csv')) {
    vehicles <- read.csv('./data/311.csv') %>%
      clean_names() %>%
      filter(grepl('Junk Vehicles', request_reason))
    
    # Convert to spatial object
    vehicles_sf <- vehicles %>%
      st_as_sf(coords = c('longitude', 'latitude'),
               crs = 4326,
               agr = 'constant',
               remove = FALSE) %>%
      na.omit()
    
    return(vehicles_sf)
  } else {
    stop("311 file not found")
  }
}

# Load chapter 66 abatements
load_abatements <- function() {
  # Check if the processed file already exists
  if (!file.exists('./data/ch66_sf.RDS')) {
    # Initial import and clean
    ch66 <- read.csv('./data/ch66.csv') %>%
      clean_names() %>%
      mutate(casefiled = as.Date(casefiled)) %>%
      mutate(year = substr(casefiled, 1, 4), .after = casefiled) %>%
      filter(year == '2023' & longitude >= -91.18) %>% # exclude cases outside of New Orleans
      filter(!is.na(longitude) & !is.na(latitude)) %>% # ensure both coordinates are non-NA
      distinct()
    
    # Convert to sf object using existing coordinates
    ch66_sf <- ch66 %>%
      st_as_sf(coords = c('longitude', 'latitude'),
               crs = 4326,
               agr = "constant") %>%
      st_make_valid()
    
    # Check if the sf object is valid
    if (any(st_is_valid(ch66_sf) == FALSE)) {
      warning("Some geometries are invalid after applying st_make_valid()")
    }
    
    # Save the processed data to an RDS file for future use
    saveRDS(ch66_sf, './data/ch66_sf.RDS')
    message("Ch 66 abatements data processed and saved.")
  } else {
    # Load the pre-processed data from the saved RDS
    ch66_sf <- readRDS('./data/ch66_sf.RDS')
    message("Ch 66 abatements data loaded from cache.")
  }
  
  return(ch66_sf)
}
```

```{r load_risk_factors}
# Load all risk factors
violations <- load_violations()
streetlights <- load_streetlights()
vehicles <- load_vehicles()
abatements <- load_abatements()
```

## 2.3 Data Aggregation by Neighborhood

```{r aggregation_functions}
# Aggregate violent calls by neighborhood
aggregate_violent_calls <- function(violent_calls, neighborhoods) {
  violent_by_neighborhood <- st_join(violent_calls, neighborhoods) %>%
    group_by(gnocdc_lab) %>%
    summarize(violent_count = n()) %>%
    st_drop_geometry()
  
  return(violent_by_neighborhood)
}

# Aggregate code violations by neighborhood
aggregate_violations <- function(violations, neighborhoods) {
  violations_by_neighborhood <- st_join(violations, neighborhoods) %>%
    group_by(gnocdc_lab) %>%
    summarize(code_violations = n()) %>%
    st_drop_geometry()
  
  return(violations_by_neighborhood)
}

# Aggregate streetlight outages by neighborhood
aggregate_streetlights <- function(lights, neighborhoods) {
  lights_by_neighborhood <- st_join(lights, neighborhoods) %>%
    group_by(gnocdc_lab) %>%
    summarize(streetlight_outages = n()) %>%
    st_drop_geometry()
  
  return(lights_by_neighborhood)
}

# Aggregate abandoned vehicles by neighborhood
aggregate_vehicles <- function(vehicles, neighborhoods) {
  vehicles_by_neighborhood <- st_join(vehicles, neighborhoods) %>%
    group_by(gnocdc_lab) %>%
    summarize(abandoned_vehicles = n()) %>%
    st_drop_geometry()
  
  return(vehicles_by_neighborhood)
}

# Aggregate chapter 66 abatements by neighborhood
aggregate_abatements <- function(ch66, neighborhoods) {
  ch66_by_neighborhood <- st_join(ch66, neighborhoods) %>%
    group_by(gnocdc_lab) %>%
    summarize(lot_abatements = n()) %>%
    st_drop_geometry()
  
  return(ch66_by_neighborhood)
}

# Join all neighborhood-level data
create_neighborhood_dataset <- function(neighborhoods, violent_by_neighborhood,
                                       violations_by_neighborhood,
                                       lights_by_neighborhood,
                                       vehicles_by_neighborhood,
                                       ch66_by_neighborhood) {
  neighborhood_data <- neighborhoods %>%
    select(gnocdc_lab, geometry) %>%
    left_join(violent_by_neighborhood, by = "gnocdc_lab") %>%
    left_join(violations_by_neighborhood, by = "gnocdc_lab") %>%
    left_join(lights_by_neighborhood, by = "gnocdc_lab") %>%
    left_join(vehicles_by_neighborhood, by = "gnocdc_lab") %>%
    left_join(ch66_by_neighborhood, by = "gnocdc_lab") %>%
    mutate(across(where(is.numeric), ~replace_na(., 0)))
  
  return(neighborhood_data)
}
```

```{r aggregate_data}
# Aggregate data by neighborhood
violent_by_neighborhood <- aggregate_violent_calls(violent_calls, neighborhoods)
violations_by_neighborhood <- aggregate_violations(violations, neighborhoods)
lights_by_neighborhood <- aggregate_streetlights(streetlights, neighborhoods)
vehicles_by_neighborhood <- aggregate_vehicles(vehicles, neighborhoods)
ch66_by_neighborhood <- aggregate_abatements(abatements, neighborhoods)

# Create final neighborhood dataset
neighborhood_data <- create_neighborhood_dataset(
  neighborhoods,
  violent_by_neighborhood,
  violations_by_neighborhood,
  lights_by_neighborhood,
  vehicles_by_neighborhood,
  ch66_by_neighborhood
)
```

## 2.4 Statistical Analysis

### 2.4.1 Correlation Analysis

```{r correlation_analysis}
# Define a function for running a correlation between risk factors and violent calls
run_correlation_analysis <- function(neighborhood_data) {
  # Prepare data for correlation
  corr_data <- neighborhood_data %>%
    st_drop_geometry() %>%
    select(violent_count, code_violations, streetlight_outages,
           abandoned_vehicles, lot_abatements)
  
  # Calculate correlation matrix
  cor_matrix <- cor(corr_data, use = "pairwise.complete.obs")
  
  # Calculate correlation with p-values
  corr_results <- data.frame()
  predictors <- c("code_violations", "streetlight_outages",
                 "abandoned_vehicles", "lot_abatements")
  
  for (predictor in predictors) {
    test <- cor.test(
      neighborhood_data[[predictor]],
      neighborhood_data[["violent_count"]],
      use = "pairwise.complete.obs"
    )
    
    corr_results <- rbind(corr_results, data.frame(
      factor = predictor,
      correlation = test$estimate,
      p_value = test$p.value,
      significance = case_when(
        test$p.value < 0.001 ~ "***",
        test$p.value < 0.01 ~ "**",
        test$p.value < 0.05 ~ "*",
        TRUE ~ "ns"
      )
    ))
  }
  
  # Return both correlation matrix and detailed test results
  return(list(
    correlation_matrix = cor_matrix,
    correlation_tests = corr_results
  ))
}

# Run the correlation analysis
correlation_results <- run_correlation_analysis(neighborhood_data)
kable(correlation_results$correlation_tests, 
      caption = "Correlation of Urban Disorder Indicators with Violent Incidents")
```

### 2.4.2 Visualizing Relationships

```{r visualize_correlations, fig.cap="Relationship between Code Violations and Violent Calls"}
# Dot plot of code violations vs violent calls with smoothing line
ggplot(neighborhood_data %>% st_drop_geometry(),
       aes(x = code_violations, y = violent_count)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "lm", color = "blue") +
  theme_solarized() +
  labs(title = "Relationship between code violations and violent calls",
       x = "Number of code violations",
       y = "Number of violent calls") +
  geom_text_repel(aes(label = str_to_title(gnocdc_lab)), size = 3, max.overlaps = 10)
```

```{r visualize_all_correlations, fig.cap="Relationship between All Risk Factors and Violent Calls"}
# First, prepare the data in long format
correlation_data <- neighborhood_data %>%
  st_drop_geometry() %>%
  select(gnocdc_lab, violent_count, code_violations, streetlight_outages,
         abandoned_vehicles, lot_abatements) %>%
  pivot_longer(cols = c(code_violations, streetlight_outages,
                         abandoned_vehicles, lot_abatements),
               names_to = "risk_factor",
               values_to = "count")

# Create plot with all risk factors colored differently
ggplot(correlation_data, aes(x = count, y = violent_count, color = risk_factor)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, aes(fill = risk_factor), alpha = 0.1) +
  scale_color_brewer(palette = "Set1",
                    labels = c("Abandoned Vehicles", "Code Violations",
                              "Lot Abatements", "Streetlight Outages")) +
  scale_fill_brewer(palette = "Set1",
                   labels = c("Abandoned Vehicles", "Code Violations",
                             "Lot Abatements", "Streetlight Outages")) +
  theme_solarized() +
  labs(title = "Relationship between risk factors and violent calls",
       x = "Count of risk factors",
       y = "Number of violent calls",
       color = "Risk factor",
       fill = "Risk factor") +
  theme(legend.position = "right")
```

# 3. Predictive Modeling

## 3.1 Data Preparation for Modeling

```{r prepare_ml_data}
prepare_ml_data <- function(neighborhood_data) {
  # Prepare data for modeling
  model_data <- neighborhood_data %>%
    st_drop_geometry() %>%
    select(violent_count, code_violations, streetlight_outages,
           abandoned_vehicles, lot_abatements) %>%
    # Remove any rows with NA values
    na.omit()
  
  # Split into training and testing sets (80/20)
  set.seed(90210)
  data_split <- initial_split(model_data, prop = 0.8)
  train_data <- training(data_split)
  test_data <- testing(data_split)
  
  # Create validation set for tuning
  set.seed(90210)
  cv_folds <- vfold_cv(train_data, v = 5)
  
  return(list(
    train_data = train_data,
    test_data = test_data,
    cv_folds = cv_folds,
    full_data = model_data
  ))
}

ml_data <- prepare_ml_data(neighborhood_data)
```

## 3.2 Linear Regression Model

```{r linear_regression}
# Re-establish seed
set.seed(90210)

# Define a function for the lm model
build_lm_model <- function(ml_data) {
  # Define the model specification
  lm_spec <- linear_reg() %>%
    set_engine("lm") %>%
    set_mode("regression")
  
  # Define the workflow
  lm_workflow <- workflow() %>%
    add_formula(violent_count ~ .) %>%
    add_model(lm_spec)
  
  # Fit the model
  lm_fit <- lm_workflow %>%
    fit(data = ml_data$train_data)
  
  # Predictions on test data
  lm_predictions <- lm_fit %>%
    predict(ml_data$test_data) %>%
    bind_cols(ml_data$test_data)
  
  # Calculate performance metrics
  lm_metrics <- lm_predictions %>%
    metrics(truth = violent_count, estimate = .pred)
  
  # Extract model coefficients
  lm_coefs <- lm_fit %>%
    extract_fit_parsnip() %>%
    tidy()
  
  return(list(
    workflow = lm_workflow,
    model = lm_fit,
    predictions = lm_predictions,
    metrics = lm_metrics,
    coefficients = lm_coefs
  ))
}

# Print results of the linear model
lm_results <- build_lm_model(ml_data)
```

```{r lm_results}
# Display linear model performance metrics
cat("Linear Model Performance Metrics:\n")
kable(lm_results$metrics)

cat("\nModel Coefficients:\n")
kable(lm_results$coefficients)
```

## 3.3 Random Forest Model

```{r random_forest}
# Re-establish seed
set.seed(90210)

# Build Random Forest model with tidymodels
build_rf_model <- function(ml_data) {
  # Define the model specification
  rf_spec <- rand_forest(
    mtry = tune(),
    trees = 1000,
    min_n = tune()
  ) %>%
    set_engine("ranger", importance = "impurity") %>%
    set_mode("regression")
  
  # Define the workflow
  rf_workflow <- workflow() %>%
    add_formula(violent_count ~ .) %>%
    add_model(rf_spec)
  
  # Define the grid for tuning
  rf_grid <- grid_regular(
    mtry(range = c(1, 4)),
    min_n(range = c(2, 10)),
    levels = 1
  )
  
  # Tune the model
  rf_tune_results <- rf_workflow %>%
    tune_grid(
      resamples = ml_data$cv_folds,
      grid = rf_grid,
      metrics = metric_set(rmse, rsq)
    )
  
  # Select best hyperparameters
    rf_best_params <- rf_tune_results %>%
    select_best(metric = "rmse")
  
  # Finalize workflow with best parameters
  rf_final_workflow <- rf_workflow %>%
    finalize_workflow(rf_best_params)
  
  # Fit the final model
  rf_final_fit <- rf_final_workflow %>%
    fit(data = ml_data$train_data)
  
  # Predictions on test data
  rf_predictions <- rf_final_fit %>%
    predict(ml_data$test_data) %>%
    bind_cols(ml_data$test_data)
  
  # Calculate performance metrics
  rf_metrics <- rf_predictions %>%
    metrics(truth = violent_count, estimate = .pred)
  
  # Extract variable importance
  rf_imp <- rf_final_fit %>%
    extract_fit_parsnip() %>%
    vip::vi()
  
  return(list(
    workflow = rf_final_workflow,
    model = rf_final_fit,
    predictions = rf_predictions,
    metrics = rf_metrics,
    importance = rf_imp,
    tuning_results = rf_tune_results
  ))
}

# Run random forest model
rf_results <- build_rf_model(ml_data)
```


```{r random_forest_results}
# Print results of the random forest model
cat("Random Forest Performance Metrics:\n")
kable(rf_results$metrics)

cat("\nVariable Importance:\n")
kable(rf_results$importance)
```

## 3.4 Elastic Net Regression Model

```{r elastic_net}
# Re-establish seed
set.seed(90210)

build_elastic_net_model <- function(ml_data) {
  # Define the model specification
  en_spec <- linear_reg(
    penalty = tune(),
    mixture = tune()
  ) %>%
    set_engine("glmnet") %>%
    set_mode("regression")
  
  # Create pre-processing recipe
  en_recipe <- recipe(violent_count ~ ., data = ml_data$train_data) %>%
    step_normalize(all_predictors())
  
  # Define the workflow
  en_workflow <- workflow() %>%
    add_recipe(en_recipe) %>%
    add_model(en_spec)
  
  # Define the grid for tuning
  # Using smaller, more appropriate values for penalty
  en_grid <- grid_regular(
    penalty(range = c(0.001, 1.0)), # Values between 0 and 1
    mixture(range = c(0, 1)), # Values between 0 and 1
    levels = c(10, 5)
  )
  
  # Tune the model
  en_tune_results <- en_workflow %>%
    tune_grid(
      resamples = ml_data$cv_folds,
      grid = en_grid,
      metrics = metric_set(rmse, rsq)
    )
  
  # Select best hyperparameters
  en_best_params <- en_tune_results %>%
    select_best(metric = "rmse")
  
  # Finalize workflow with best parameters
  en_final_workflow <- en_workflow %>%
    finalize_workflow(en_best_params)
  
  # Fit the final model
  en_final_fit <- en_final_workflow %>%
    fit(data = ml_data$train_data)
  
  # Predictions on test data
  en_predictions <- en_final_fit %>%
    predict(ml_data$test_data) %>%
    bind_cols(ml_data$test_data)
  
  # Calculate performance metrics
  en_metrics <- en_predictions %>%
    metrics(truth = violent_count, estimate = .pred)
  
  # Extract model coefficients - handle potential error
  en_coefs <- tryCatch({
    en_final_fit %>%
      extract_fit_parsnip() %>%
      tidy()
  }, error = function(e) {
    message("Could not extract coefficients: ", e$message)
    return(data.frame(term = character(), estimate = numeric()))
  })
  
  return(list(
    workflow = en_final_workflow,
    model = en_final_fit,
    predictions = en_predictions,
    metrics = en_metrics,
    coefficients = en_coefs,
    tuning_results = en_tune_results
  ))
}

# Run the elastic net model
elastic_net_results <- build_elastic_net_model(ml_data)
```

```{r elastic_net_results}
# Print results of the elastic net model
cat("Elastic Net Model Performance Metrics:\n")
kable(elastic_net_results$metrics)

cat("\nElastic Net Coefficients:\n")
tryCatch({
  kable(elastic_net_results$coefficients)
}, error = function(e) {
  cat("Could not display coefficients:", e$message)
})
```

## 3.5 Multivariate Adaptive Regression Splines (MARS)

```{r mars}
# Re-establish seed
set.seed(90210)

build_mars_model <- function(ml_data) {
  # Define the model specification
  mars_spec <- mars(
    num_terms = tune(),
    prod_degree = tune()
  ) %>%
    set_engine("earth") %>%
    set_mode("regression")
  
  # Define the workflow
  mars_workflow <- workflow() %>%
    add_formula(violent_count ~ .) %>%
    add_model(mars_spec)
  
  # Define the grid for tuning
  mars_grid <- grid_regular(
    num_terms(range = c(2, 8)), # Number of terms to retain
    prod_degree(range = c(1, 2)), # Product degree (1 = additive, 2 = two-way interactions)
    levels = c(4, 2)
  )
  
  # Tune the model
  mars_tune_results <- mars_workflow %>%
    tune_grid(
      resamples = ml_data$cv_folds,
      grid = mars_grid,
      metrics = metric_set(rmse, rsq)
    )
  
  # Select best hyperparameters
  mars_best_params <- mars_tune_results %>%
    select_best(metric = "rmse")
  
  # Finalize workflow with best parameters
  mars_final_workflow <- mars_workflow %>%
    finalize_workflow(mars_best_params)
  
  # Fit the final model
  mars_final_fit <- mars_final_workflow %>%
    fit(data = ml_data$train_data)
  
  # Predictions on test data
  mars_predictions <- mars_final_fit %>%
    predict(ml_data$test_data) %>%
    bind_cols(ml_data$test_data)
  
  # Calculate performance metrics
  mars_metrics <- mars_predictions %>%
    metrics(truth = violent_count, estimate = .pred)
  
  return(list(
    workflow = mars_final_workflow,
    model = mars_final_fit,
    predictions = mars_predictions,
    metrics = mars_metrics,
    tuning_results = mars_tune_results
  ))
}

# Run the MARS model
mars_results <- build_mars_model(ml_data)
```

```{r mars_results}
# Print results of the MARS model
cat("MARS Performance Metrics:\n")
kable(mars_results$metrics)
```

## 3.6 Model Comparison

```{r compare_models}
compare_models <- function(lm_results, rf_results, elastic_net_results, mars_results) {
  # Create comparison dataframe
  model_names <- c("Linear Regression", "Random Forest", "Elastic Net", "MARS")
  
  # Prepare metrics for comparison
  get_rmse <- function(result) {
    result$metrics %>%
      filter(.metric == "rmse") %>%
      pull(.estimate)
  }
  
  get_rsq <- function(result) {
    result$metrics %>%
      filter(.metric == "rsq") %>%
      pull(.estimate)
  }
  
  # Create comparison dataframe
  rmse_values <- c(
    get_rmse(lm_results),
    get_rmse(rf_results),
    get_rmse(elastic_net_results),
    get_rmse(mars_results)
  )
  
  rsq_values <- c(
    get_rsq(lm_results),
    get_rsq(rf_results),
    get_rsq(elastic_net_results),
    get_rsq(mars_results)
  )
  
  # Create and return the comparison dataframe
  comparison_df <- data.frame(
    Model = model_names,
    RMSE = round(rmse_values, 2),
    R_squared = round(rsq_values, 2)
  )
  
  return(comparison_df)
}

# Run function to compare all models
comparison_results <- compare_models(
  lm_results,
  rf_results,
  elastic_net_results,
  mars_results
)

# Sort by RMSE (ascending)
comparison_results %>%
  arrange(RMSE) %>%
  kable(caption = "Comprehensive Model Performance Comparison")
```


# 4. Results and Discussion

## 4.1 Correlation Analysis Findings

Our correlation analysis revealed strong relationships between several urban disorder indicators and violent incidents in New Orleans neighborhoods. Code violations showed the strongest correlation with violent incidents (r = 0.74, p < 0.001), followed by streetlight outages (r = 0.69, p < 0.001) and abandoned vehicles (r = 0.65, p < 0.001). Lot abatements showed a weaker and non-significant correlation (r = 0.16, p = 0.18).

These findings align with the "broken windows" theory of urban disorder, which suggests that visible signs of disorder and neglect may contribute to an environment where crime is more prevalent. The spatial visualization of violent incidents shows clustering in certain neighborhoods, which corresponds with areas that have higher concentrations of code violations and other disorder indicators.

## 4.2 Predictive Model Performance

We compared four predictive models to understand the relationships between urban disorder indicators and violent incidents: linear regression, random forest, elastic net regression, and multivariate adaptive regression splines (MARS).

### 4.2.1 Linear Regression

The linear regression model achieved an RMSE of 314.46 and an R-squared of 0.72, indicating it explains approximately 72% of the variance in violent incidents across neighborhoods. The model coefficients reveal that code violations have the strongest predictive relationship with violent incidents, with each additional code violation associated with approximately 0.90 more violent incidents, holding other variables constant. Streetlight outages were also significant predictors, with each outage associated with about 0.31 additional violent incidents. Abandoned vehicles and lot abatements showed minimal and non-significant associations in the linear model.

### 4.2.2 Random Forest

The random forest model performed comparably to the linear regression model, with an RMSE of 322.87 and a slightly higher R-squared of 0.73. This model's performance suggests that the relationships between the predictors and violent incidents are predominantly linear, as the more flexible random forest model did not substantially outperform the linear model. The random forest variable importance metrics identified code violations as the most important predictor, followed by streetlight outages, abandoned vehicles, and lot abatements, in that order. This consistency with the linear model strengthens our confidence in the importance of code violations as a predictor of neighborhood violence.

### 4.2.3 Elastic Net Regression

The elastic net model, which combines L1 and L2 regularization to handle potential multicollinearity among predictors, achieved an RMSE of approximately 329.14 and an R-squared of 0.70. Though slightly underperforming compared to the linear regression and random forest models, the elastic net's performance suggests that multicollinearity among our predictors is not a major concern in this dataset. The regularization path identified similar important predictors as the other models, with code violations and streetlight outages having the strongest non-zero coefficients.

### 4.2.4 Multivariate Adaptive Regression Splines (MARS)

The MARS model, which can capture non-linear relationships between predictors and the outcome, achieved an RMSE of 318.21 and an R-squared of 0.73. This performance is comparable to the random forest model and slightly better than the elastic net. The MARS model's ability to identify potential non-linearities in the relationships between predictors and violent incidents provides additional insights. The model identified similar key predictors to the other approaches, suggesting that while there may be some non-linearity in the relationships, the overall pattern of importance among predictors remains consistent.

### 4.2.5 Model Comparison

Comparing all four models, the linear regression model achieved the lowest RMSE (314.45), suggesting it provides the most accurate predictions on our test data. The random forest and MARS models had slightly higher R-squared values (0.73), indicating they explained marginally more variance than the linear regression (0.72) and elastic net (0.70) models. The consistency in performance across different modeling approaches suggests that our findings regarding the importance of code violations and streetlight outages as predictors of violent incidents are robust.

The minimal performance differences between linear and more complex models suggest that the relationships between urban disorder indicators and violent incidents are predominantly linear in nature. This is advantageous from a policy perspective, as linear relationships are more straightforward to interpret and communicate to stakeholders.

## 4.3 Spatial Patterns

The spatial distribution of violent incidents across New Orleans neighborhoods reveals distinct patterns. Certain neighborhoods consistently show higher concentrations of both violent incidents and urban disorder indicators. This spatial clustering suggests that interventions focused on specific high-risk neighborhoods might be more effective than city-wide approaches.

The neighborhoods with the highest concentrations of violent incidents also tend to have the highest numbers of code violations and streetlight outages, further supporting the correlation between these factors. This spatial coincidence of urban disorder and violence provides additional evidence for the theoretical connection between physical disorder and crime.

## 4.4 Implications of Findings

These findings have several important implications:

1. **Predictive Value of Code Violations**: The strong and consistent relationship between code violations and violent incidents across all models suggests that code violations may serve as a valuable early indicator of neighborhood conditions conducive to violence. This provides empirical support for policy approaches that target code enforcement as part of comprehensive violence reduction strategies.

2. **Importance of Infrastructure Maintenance**: The significant relationship between streetlight outages and violent incidents highlights the importance of basic infrastructure maintenance in crime prevention. This finding aligns with situational crime prevention theory, which emphasizes how environmental conditions can influence criminal opportunities (Braga, 2024).

3. **Targeted Intervention Opportunities**: The spatial clustering of both disorder indicators and violent incidents suggests that targeted interventions in high-risk neighborhoods could be particularly effective. Resources for both code enforcement and crime prevention might be most efficiently allocated by focusing on neighborhoods with high concentrations of code violations and streetlight outages.

4. **Integrated Approach**: The multiple predictors identified suggest that an integrated approach addressing various forms of physical disorder simultaneously might be more effective than focusing on single factors in isolation.

# 5. Conclusion

## 5.1 Summary of Findings

This analysis identified significant relationships between urban disorder indicators and violent incidents in New Orleans neighborhoods. Code violations emerged as the strongest predictor of neighborhood violence, followed by streetlight outages and abandoned vehicles. Our predictive models achieved reasonable accuracy, with R-squared values around 0.72-0.73, indicating that these disorder indicators explain a substantial portion of the variance in violent incidents across neighborhoods.

## 5.2 Implications for Policy and Practice

These findings have several implications for public safety initiatives in New Orleans:

1. **Targeted Code Enforcement**: Given the strong relationship between code violations and violent incidents, strengthening code enforcement in high-risk neighborhoods could potentially contribute to violence reduction.

2. **Infrastructure Maintenance**: The significant association between streetlight outages and violent incidents suggests that improving street lighting maintenance could be a relatively straightforward intervention to enhance neighborhood safety.

3. **Data-Driven Resource Allocation**: The spatial patterns identified in this analysis can inform more targeted allocation of public safety resources to neighborhoods with higher concentrations of risk factors.

4. **Cross-Department Collaboration**: Addressing neighborhood violence effectively may require collaboration across multiple city departments, including code enforcement and public works.

## 5.3 Limitations

Several limitations should be considered when interpreting these results:

1. **Correlation vs. Causation**: While our analysis identified strong correlations between disorder indicators and violent incidents, we cannot establish causation. Other unmeasured factors might influence both disorder indicators and violent incidents.

2. **Data Quality**: Our analysis depends on reported incidents, which may not capture all relevant events. Underreporting may vary across neighborhoods, as calls for service suffer drastically from reporting bias, thus limiting the predictive capability of measuring crime accordingly (Buil-Gil, Moretti & Langton, 2021).

3. **Temporal Dynamics**: This analysis used data from 2023 only, providing a snapshot rather than capturing trends over time.

4. **Socioeconomic Factors**: Our models did not include socioeconomic variables like poverty rates, unemployment, or educational attainment, which may be important factors in understanding neighborhood violence.

## 5.4 Future Research

Future research could build on this work in several ways:

1. **Longitudinal Analysis**: Examining how changes in disorder indicators over time relate to changes in violent incidents could provide stronger evidence for causal relationships.

2. **Additional Variables**: Incorporating socioeconomic data, housing characteristics, and other neighborhood attributes could improve model performance and provide a more comprehensive understanding of neighborhood violence.

3. **Intervention Studies**: Evaluating the impact of targeted interventions (e.g., enhanced code enforcement or streetlight maintenance) on violent incidents would provide valuable evidence on effective strategies.

4. **Machine Learning Approaches**: More advanced machine learning techniques could potentially identify complex non-linear relationships and interaction effects among predictors.

In conclusion, this analysis provides evidence that easily observable and measurable indicators of urban disorder have predictive value for understanding patterns of violent incidents in New Orleans neighborhoods. These findings can inform more targeted and potentially more effective approaches to improving public safety.

# 6. References

Braga, Anthony A. (2024). Disorder policing to reduce crime: An updated systematic review and meta-analysis. *Criminology & Public Policy*, 23(3), 745-775. https://doi.org/10.1111/1745-9133.12667

Buil-Gil, D., Moretti, A. & Langton, S.H. (2021). The accuracy of crime statistics: assessing the impact of police data bias on geographic crime analysis. *Journal of Experimental Criminology*, 18, 515–541. https://doi.org/10.1007/s11292-021-09457-y

City of New Orleans. (2023). Open Data Portal. Retrieved from https://data.nola.gov/

Kelling, G. L., & Wilson, J. Q. (1982). Broken windows: The police and neighborhood safety. *Atlantic Monthly*, 249(3), 29-38. https://www.theatlantic.com/magazine/archive/1982/03/broken-windows/304465/

